\message{ !name(sphere_generator.tex)}\documentclass{article}

\title{Sphere Generator: Techniques}
\author{Jonah Miller\\\textit{jonah.maxwell.miller@gmail.com}

\begin{document}

\message{ !name(sphere_generator.tex) !offset(-3) }


\maketitle

This is documentation for sphere_generator.py, a program to generate
spheres of arbitrary surface area triangulated by equilateral
triangles. The sphere is represented as a list of triples, each triple
containing 3 vertex id numbers. Matching up the IDs yields an object
homeomorphic to a sphere, and with an (ideally) uniform curvature.

This is a description of the ideas used to make the sphere generator
program. The README and the programmer's guide offer more details. In
this document I will describe how the program selects for standard
deviation surface area of a given sphere.

Associated with each vertex is a point curvature:
$$K_p(v) = 2(2 - \theta_t N_T(v)),$$
where $\theta_t = \pi/3$ is the interior angle of an equilateral
triangle and $N_t(v)$ is the total number of triangles connected to a
given vertex. This description of curvature is given by Regge
calculus.

The mean curvature is then 
$$K = \frac{1}{N_v}\sum_{vertices, \ v} K_p(v),$$
where $N_v$ is the total number of vertices in the sphere.

The standard deviation of curvature, $\sigma$ is then given by
$$\sigma^2 = \frac{1}{N_v} \sum_{vertices, \ v}\left(K - K_p(v)\right)^2.$$

Let $\epsilon_a$ and $\epsilon_\sigma$ be numbers between 0 and 1. Let
the target standard deviation---i.e., the standard deviation we want
the sphere to have---be $\sigma_t$ and the target surface area of the
sphere---i.e., the surface area we want the sphere to have---be
$A_t$. The \textit{fitness function} for a given sphere $S$ is
$$f(S) = e^{-\epsilon_a (A - A_t)}e^{-\epsilon_\sigma (\sigma - \sigma_t)},$$
where $A$ is the surface area of the sphere at a given time. 

The metropolis algorithm does the following:



\end{document}
\message{ !name(sphere_generator.tex) !offset(-51) }
